A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-02-01 doc/large_strain.tex
Move comment related to implementation details of GMG
Fiddle with intro to numerical examples; expanded on some points

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number $N$ of unknown DoFs is desirable.
The efficiency of standard iterative solvers and preconditioners deteriorates for large systems of linear equations due to increasing iteration counts under mesh refinement.
In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{deformed.png}
      \caption{2D deformed mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{deformed_3d.png}
    \caption{3D deformed mesh}
  \end{subfigure}
  \caption{Deformed meshes at the final loading step with quadratic elements and two global mesh refinements.}%
  \label{fig:miehe_deformed}
\end{figure}

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
%
\changeJP{The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.}
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements \changeJP{for the mesh introduced in figure \ref{fig:miehe_fine_level}}.
{\color{red}
Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}
\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
  }
  \label{tab:input_parameters_2d}
\end{table}

@@

{\color{red}In this study, we limit ourselves to a comparison on a node level, which should already indicate the differences between the matrix-free and matrix-based approaches.}
The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
{\color{red}
We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach.
On the other hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
}
%
The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.

Computations \changeJP{were performed} on a single node of two clusters:
A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. fully occupied node.}

>>>>

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number $N$ of unknown DoFs is desirable.
The efficiency of standard iterative solvers and preconditioners deteriorates for large systems of linear equations due to increasing iteration counts under mesh refinement.
In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
\changeJP{Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
%
Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{deformed.png}
      \caption{2D deformed mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{deformed_3d.png}
    \caption{3D deformed mesh}
  \end{subfigure}
  \caption{Deformed meshes at the final loading step with quadratic elements and two global mesh refinements.}%
  \label{fig:miehe_deformed}
\end{figure}
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
%
\changeJP{The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.}
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements \changeJP{for the mesh introduced in figure \ref{fig:miehe_fine_level}}.
%
\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
  }
  \label{tab:input_parameters_2d}
\end{table}

@@

\changeJP{%
Aligned with the approach taken in similar previous investigations \cite{}\todo{Cite Martin's papers?}, in this study we limit ourselves to comparisons made on a node level.
Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange though interconnects via MPI.
}
The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
{\color{red}
We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach \changeJP{with a high polynomial order finite element}.
\changeJP{Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound}.
}

%
% Below are are implementational details that might affect the comparative result
%
The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

Computations \changeJP{were performed} on a single node of two clusters:
A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. \changeJP{one} fully occupied node.}
